---
title: "Two Parameter Logistic Model"
author: "Erick A. Chacon-Montalvan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Two Parameter Logistic Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")
knitr::opts_chunk$set(fig.width = 6, fig.height = 4)
knitr::opts_chunk$set(comment = "#>")
options(width = 100)
```

In this vignette, we show how to use the **spmirt** package to fit a 2 parameter
logistic model used in item response theory. Let $Y_{ij}$ be the response for item
$j$ in the individual $i$. The model can be defined by using an auxiliary variable
$Z_{ij}$ such as
\begin{align}
  {Y}_{ij}  & =
  \left\lbrace
  \begin{array}[2]{cc}
    1 & \text{if} ~ {Z}_{ij} > 0\\
    0 & \text{otherwise}
  \end{array}
  \right.\\
  {Z}_{ij} & = c_j + a_j\theta_i + \epsilon_{ij},
  ~~ \epsilon_{ij} \sim {N}(0, 1)\\
  {\theta}_i & \sim {N}({0}, {I}_m)\\
  c_j & \sim {N}(0, \sigma_c^2)\\
  {a}_j & \sim {N}(0, \sigma_a^2{I}_m)\\
\end{align}

### Required packages

```{r}
rm(list = ls())
library(spmirt)
library(datasim)
library(coda)
library(tidyverse)
```

### Simulation of the data

```{r}
q <- 10
n <- 300
difficulty <- matrix((1:q - 5)/10 * 2, nrow = 1)
discrimination <- matrix(seq(0.4, 1.5, length.out = q), nrow = 1)
discrimination[1,1] = 2

f <- list(
  prob ~ mfa(ones, beta = get("difficulty")) +
    re(id, sigma = 1, q = get("q")):mfa(ones, beta = get("discrimination")),
  size ~ I(1)
  )

data_long <- sim_model(formula = f,
                        link_inv = list(pnorm, identity),
                        generator = rbinom,
                        responses = q,
                        n = n,
                        seed = 3
                        )

data_long <- dplyr::rename(data_long, subject = id, ability = re.id.prob,
                           item = response_label, y = response)
```

```{r, echo = FALSE, results = 'asis'}
knitr::kable(head(data_long, 20))
```

### Exploratory analysis

We can see the relationship between the latent ability and the proportion of items
endorsed.

```{r}
explor <- data_long %>%
  group_by(subject) %>%
  summarize(endorse = mean(y), ability = unique(ability))
ggplot(explor, aes(ability, endorse)) + geom_point(alpha = 0.5)

```


### Fitting the model

```{r}
iter <- 1000
system.time(samples <- ifa_gibbs(data_long$y, n, q, iter))
purrr::map(samples, dim)
samples$firstA
samples$firstA <- NULL
plot(samples$a[, 1:2], pch = 19, col = rgb(0, 0, 0, 0.2))
points(-discrimination[1], -discrimination[2], pch = 19, col = 2)
points(discrimination[1], discrimination[2], pch = 19, col = 2)
```

## Evaluate the results

```{r}
#
# # Organize and summarise output
# samples_after_burin <- samples %>% purrr::map(~ .[(0.5 * iter):iter, ])
# samples_params <- samples_after_burin[1:3]
# real_params <- list(theta = unique(data_long$ability),
#                     c = as.numeric(difficulty),
#                     a = as.numeric(discrimination))
# samples_params_summary <- samples_params %>%
#   map(~ apply(., 2, function (x) quantile(x, c(0.025, 0.5, 0.975)))) %>%
#   map(~ as_tibble(t(.))) %>%
#   map(~ setNames(., make.names(names(.)))) %>%
#   map2(real_params, ~ mutate(.x, param = .y)) %>%
#   setNames(c("Latent ability", "Item difficulty", "Item discrimination"))
#
# # Visualize convergence
# opar <- par(mfrow = c(2, 2), mar = c(2, 1.9, 2, 0.6))
# # layout(matrix(1:2))
# map(samples_after_burin, ~ plot(mcmc(.[, 1:5]), auto.layout = FALSE))
# par(opar)
#
# # Visualize results
# samples_params_summary %>%
#   map(~ ggplot(., aes(X50., param)) +
#           geom_errorbar(aes(ymin = X2.5., ymax = X97.5.)) +
#           geom_point(col = 2)
#         )
```
