---
title: "Multidimensional Two Parameter Logistic Model"
author: "Erick A. Chacon-Montalvan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Two Parameter Logistic Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")
knitr::opts_chunk$set(fig.width = 6, fig.height = 4)
# knitr::opts_chunk$set(fig.width = 6, fig.height = 4)
knitr::opts_chunk$set(comment = "#>")
options(width = 100)
```

In this vignette, we show how to use the **spmirt** package to fit a
multidimensional 2 parameter logistic model used in item response theory. Let
$Y_{ij}$ be the response for item
$j$ in the individual $i$. The model can be defined by using an auxiliary variable
$Z_{ij}$ such as
\begin{align}
  {Y}_{ij}  & =
  \left\lbrace
  \begin{array}[2]{cc}
    1 & \text{if} ~ {Z}_{ij} > 0\\
    0 & \text{otherwise}
  \end{array}
  \right.\\
  {Z}_{ij} & = c_j + a_j^\intercal\theta_i + \epsilon_{ij},
  ~~ \epsilon_{ij} \sim {N}(0, 1)\\
  {\theta}_i & \sim {N}({0}, {I}_m)\\
  c_j & \sim {N}(0, \sigma_c^2)\\
  {a}_j & \sim {N}(0, \sigma_a^2{I}_m)\\
\end{align}

### Required packages

```{r}
rm(list = ls())
library(spmirt)
library(datasim)
library(tidyverse)
```

### Simulation of the data

```{r}
q <- q1 <- q2 <- 10
n <- 300
difficulty <- matrix((1:q - 5)/10 * 2, nrow = 1)
discrimination1 <- matrix(seq(0.4, 1.5, length.out = q), nrow = 1)
discrimination2 <- matrix(runif(q, 0, 2), nrow = 1)
discrimination1[1,1] = 1
discrimination2[1,2] = 1
discrimination2[1,1] <- 0

f <- list(
  prob ~ mfa(ones, beta = get("difficulty")) +
    re(id, sigma = 1, q = get("q1")):mfa(ones, beta = get("discrimination1")) +
    re(id, sigma = 1, q = get("q2")):mfa(ones, beta = get("discrimination2")),
  size ~ I(1)
  )

data_long <- sim_model(formula = f,
                        link_inv = list(pnorm, identity),
                        generator = rbinom,
                        responses = q,
                        n = n
                        # ,
                        # seed = 2
                        # seed = 3
                        )

data_long <- dplyr::rename(data_long, subject = id, ability1 = re.id.prob,
                           ability2 = re.id.prob.1,
                           item = response_label, y = response)
```

```{r, echo = FALSE, results = 'asis'}
knitr::kable(head(data_long, 20))
```

### Exploratory analysis

We can see the relationship between the latent abilities and the proportion of
endorsed items.

```{r}
explor <- data_long %>%
  group_by(subject) %>%
  summarize(endorse = mean(y),
            ability1 = unique(ability1),
            ability2 = unique(ability2))
ggplot(explor, aes(ability1, endorse)) + geom_point(alpha = 0.5)
ggplot(explor, aes(ability2, endorse)) + geom_point(alpha = 0.5)
```


## Fitting the model

```{r}
iter <- 1000
system.time(samples <- ifa_gibbs(data_long$y, n, q, iter, 2))
purrr::map(samples, dim)
```

### Check sampling

Sampling of the first two discrimination parameters.

```{r}

gg_2scatter <- function (samples, real) {

  data <- samples %>%
    as_tibble() %>%
    setNames(paste0(rep(c("x-", "y-"), each = ncol(samples)/2),
                    rep(1:(ncol(samples)/2), 2))) %>%
    mutate(iteration = 1:n()) %>%
    gather(varname, varvalue, -iteration) %>%
    tidyr::extract(varname, c("var", "number"), "([[:alpha:]])-([0-9]+)",
                   convert = TRUE) %>%
    spread(var, varvalue) %>%
    group_by(number) %>%
    nest()
  real <- real %>%
    as_tibble() %>%
    mutate(id = 1:n()) %>%
    group_by(id) %>%
    nest()
  data$data %>%
      map2(real$data, ~ ggplot(.x, aes(x, y)) +
      geom_point(alpha = 0.5) +
      geom_path(alpha = 0.4, linetype = 2) +
      geom_point(aes(V1, V2), data = .y, col = 2))
}
gg_2scatter(samples$a, t(rbind(discrimination1, discrimination2)))

```

### Remove burn-in and summarise parameters

```{r}

# Organize and summarise output
samples_after_burin <- samples %>% purrr::map(~ .[(0.6 * iter):iter, ])
samples_params <- samples_after_burin[1:3]
samples_params_summary <- samples_params %>%
  map(~ apply(., 2, function (x) quantile(x, c(0.025, 0.5, 0.975)))) %>%
  map(~ as_tibble(t(.))) %>%
  map(~ setNames(., make.names(names(.))))

```

### Visualize trace-plots

```{r}
gg_trace <- function (samples, name) {
  samples %>%
    as_tibble() %>%
    setNames(paste0(name, 1:ncol(samples))) %>%
    mutate(iteration = 1:n()) %>%
    gather(varname, varvalue, -iteration) %>%
    ggplot(aes(iteration, varvalue, group = varname, col = varname)) +
      geom_path(alpha = 0.4, linetype = 1)
}

samples_params %>%
  map2(c("theta", "c", "a"), ~ gg_trace(.x[,1:5], .y))

```


```{r}

# Matching parameters
samples_params_summary$theta <- samples_params_summary$theta %>%
  mutate(
    param1 = c(unique(data_long$ability1), unique(data_long$ability2)),
    param2 = c(unique(data_long$ability1), - unique(data_long$ability2)),
    param3 = c(- unique(data_long$ability1), unique(data_long$ability2)),
    param4 = - c(unique(data_long$ability1), unique(data_long$ability2)),
    param5 = c(unique(data_long$ability2), unique(data_long$ability1)),
    param6 = c(unique(data_long$ability2), - unique(data_long$ability1)),
    param7 = c(- unique(data_long$ability2), unique(data_long$ability1)),
    param8 = - c(unique(data_long$ability2), unique(data_long$ability1)),
         )
samples_params_summary$a <- samples_params_summary$a %>%
  mutate(
    param1 = c(discrimination1, discrimination2),
    param2 = c(discrimination1, - discrimination2),
    param3 = c(- discrimination1, discrimination2),
    param4 = - c(discrimination1, discrimination2),
    param5 = c(discrimination2, discrimination1),
    param6 = c(discrimination2, - discrimination1),
    param7 = c(- discrimination2, discrimination1),
    param8 = - c(discrimination2, discrimination1),
         )
samples_params_summary$c$param <- c(difficulty)

# Visualize results
samples_params_summary$theta %>% gather(key, param, param1:param4) %>%
  ggplot(., aes(X50., param)) +
    geom_errorbar(aes(ymin = X2.5., ymax = X97.5.)) +
    geom_point(col = 2) +
    facet_wrap(~ key)
samples_params_summary$theta %>% gather(key, param, param5:param8) %>%
  ggplot(., aes(X50., param)) +
    geom_errorbar(aes(ymin = X2.5., ymax = X97.5.)) +
    geom_point(col = 2) +
    facet_wrap(~ key)

samples_params_summary$a %>% gather(key, param, param1:param4) %>%
  ggplot(., aes(X50., param)) +
    geom_errorbar(aes(ymin = X2.5., ymax = X97.5.)) +
    geom_point(col = 2) +
    facet_wrap(~ key)
samples_params_summary$a %>% gather(key, param, param5:param8) %>%
  ggplot(., aes(X50., param)) +
    geom_errorbar(aes(ymin = X2.5., ymax = X97.5.)) +
    geom_point(col = 2) +
    facet_wrap(~ key)


samples_params_summary$c %>%
  ggplot(., aes(X50., param)) +
    geom_errorbar(aes(ymin = X2.5., ymax = X97.5.)) +
    geom_point(col = 2)


```

Correlation

```{r}
# ability_real <- matrix(unique(c(data_long$ability1, data_long$ability2)), ncol = 2)
# ability_estimate <- matrix(samples_params_summary$theta$X50., ncol = 2)
# cor(ability_real)
# cor(ability_estimate)
# cov(ability_real)
# cov(ability_estimate)
#
# A <- cbind(c(discrimination1), c(discrimination2))
# solve(crossprod(A) + diag(1, 2))
#
# tcrossprod(A)
#
# A_est <- matrix(samples_params_summary$a$X50., ncol = 2)
#
# plot(tcrossprod(A), tcrossprod(A_est))
# abline(0, 1)


```
