---
title: "Two Parameter Logistic Model"
author: "Erick A. Chacon-Montalvan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Two Parameter Logistic Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")
knitr::opts_chunk$set(fig.width = 6, fig.height = 4)
knitr::opts_chunk$set(comment = "#>")
options(width = 100)
```

In this vignette, we show how to use the **spmirt** package to fit a 2 parameter
logistic model used in item response theory. Let $Y_{ij}$ be the response for item
$j$ in the individual $i$. The model can be defined by using an auxiliary variable
$Z_{ij}$ such as
\begin{align}
  {Y}_{ij}  & =
  \left\lbrace
  \begin{array}[2]{cc}
    1 & \text{if} ~ {Z}_{ij} > 0\\
    0 & \text{otherwise}
  \end{array}
  \right.\\
  {Z}_{ij} & = c_j + a_j\theta_i + \epsilon_{ij},
  ~~ \epsilon_{ij} \sim {N}(0, 1)\\
  {\theta}_i & \sim {N}({0}, {I}_m)\\
  c_j & \sim {N}(0, \sigma_c^2)\\
  {a}_j & \sim {N}(0, \sigma_a^2{I}_m)\\
\end{align}

### Required packages

```{r}
rm(list = ls())
library(spmirt)
library(datasim)
library(tidyverse)
```

### Simulation of the data

```{r}
q <- 10
n <- 300
difficulty <- matrix((1:q - 5)/10 * 2, nrow = 1)
discrimination <- matrix(seq(0.4, 1.5, length.out = q), nrow = 1)
discrimination[1,1] = 2

f <- list(
  prob ~ mfa(ones, beta = get("difficulty")) +
    re(id, sigma = 1, q = get("q")):mfa(ones, beta = get("discrimination")),
  size ~ I(1)
  )

data_long <- sim_model(formula = f,
                        link_inv = list(pnorm, identity),
                        generator = rbinom,
                        responses = q,
                        n = n,
                        seed = 3)

data_long <- dplyr::rename(data_long, subject = id, ability = re.id.prob,
                           item = response_label, y = response)
```

```{r, echo = FALSE, results = 'asis'}
knitr::kable(head(data_long, 20))
```

### Exploratory analysis

We can see the relationship between the latent ability and the proportion of items
endorsed.

```{r}
explor <- data_long %>%
  group_by(subject) %>%
  summarize(endorse = mean(y), ability = unique(ability))
ggplot(explor, aes(ability, endorse)) + geom_point(alpha = 0.5)

```


### Fitting the model

```{r}
iter <- 1000
system.time(samples <- ifa_gibbs(data_long$y, n, q, iter))
purrr::map(samples, dim)
```

### Check sampling

Sampling of the first two discrimination parameters.

```{r}
discr_df <- data.frame(a1 = discrimination[1], a2 = discrimination[2])
samples$a[, 1:2] %>%
  as_tibble() %>%
  setNames(c("a1", "a2")) %>%
  ggplot(aes(a1, a2)) +
    geom_point(alpha = 0.5) +
    geom_path(alpha = 0.4, linetype = 2) +
    geom_point(data = discr_df, col = 2)
```

### Remove burn-in and summarise parameters

```{r}

# Organize and summarise output
samples_after_burin <- samples %>% purrr::map(~ .[(0.5 * iter):iter, ])
samples_params <- samples_after_burin[1:3]
real_params <- list(theta = unique(data_long$ability),
                    c = as.numeric(difficulty),
                    a = as.numeric(discrimination))
samples_params_summary <- samples_params %>%
  map(~ apply(., 2, function (x) quantile(x, c(0.025, 0.5, 0.975)))) %>%
  map(~ as_tibble(t(.))) %>%
  map(~ setNames(., make.names(names(.)))) %>%
  map2(real_params, ~ mutate(.x, param = .y)) %>%
  setNames(c("Latent ability", "Item difficulty", "Item discrimination"))

```

### Visualize trace-plots

```{r}
gg_trace <- function (samples, name) {
  samples %>%
    as_tibble() %>%
    setNames(paste0(name, 1:ncol(samples))) %>%
    mutate(iteration = 1:n()) %>%
    gather(varname, varvalue, -iteration) %>%
    ggplot(aes(iteration, varvalue, group = varname, col = varname)) +
      geom_path(alpha = 0.4, linetype = 1)
}

samples_params %>%
  map2(c("theta", "c", "a"), ~ gg_trace(.x[,1:5], .y))

```

### Visualize credible intervals

```{r}
# Visualize results
samples_params_summary %>%
  map(~ ggplot(., aes(X50., param)) +
          geom_errorbar(aes(ymin = X2.5., ymax = X97.5.), width = 0.05) +
          geom_point(col = 2)
        )
```
