---
title: "Spatial Item Factor Analysis"
author: "Erick A. Chacon-Montalvan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatial Item Factor Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")
knitr::opts_chunk$set(fig.width = 7, fig.height = 4)
knitr::opts_chunk$set(comment = "#>")
options(width = 100)
```

In this vignette, we show how to use the **spifa** package to fit a
multidimensional 2 parameter logistic model used in item response theory. Let
$Y_{ij}$ be the response for item
$j$ in the individual $i$. The model can be defined by using an auxiliary variable
$Z_{ij}$ such as
\begin{align}
  {Y}_{ij}  & =
  \left\lbrace
  \begin{array}[2]{cc}
    1 & \text{if} ~ {Z}_{ij} > 0\\
    0 & \text{otherwise}
  \end{array}
  \right.\\
  {Z}_{ij} & = c_j + a_j^\intercal\theta_i + \epsilon_{ij},
  ~~ \epsilon_{ij} \sim {N}(0, 1)\\
  {\theta}_i & \sim {N}({0}, {I}_m)\\
  c_j & \sim {N}(0, \sigma_c^2)\\
  {a}_j & \sim {N}(0, \sigma_a^2{I}_m)\\
\end{align}

### Required packages

```{r}
rm(list = ls())
library(datasim)
library(tidyverse)
library(spmirt)
```

### Simulation of the data

First, simulate spatially correlated factors:

```{r}
Corr <- matrix(c(1, -0.3, 0, -0.3, 1, 0.3, 0, 0.3, 1), nrow = 3)
sigmas <- rep(0.4^0.5, 3)
D <- diag(sigmas)
Cov <- D %*% Corr %*% D

# beta <- c(-0.5, 0, 0.5)
beta <- c(0, 0, 0)
variance <- 0.6 * matrix(c(1, 0, 0, 0, 1, 0, 0, 0, 1), nrow = 3)
cor.model <- "exp_cor"
cor.params <- list(list(phi = 0.2), list(phi = 0.2), list(phi = 0.2))

f <- list(
  mean ~ mfe(x1, beta = get("beta")) +
    mre(factor(id), sigma = get("Cov")) +
    mgp(list(s1, s2), variance = get("variance"), cor.model = get("cor.model"),
        cor.params = get("cor.params"), range = 2),
  sd ~ I(0)
  )

n <- 300
m <- 3
data_geo <- sim_model(formula = f, n = n, responses = m)
# knitr::kable(head(data_model, 10))

```

First, we define the parameters for the item response model.

#### VISUALIZE MULTIVARIATE SPATIAL DATA

```{r}
ggplot(data_geo, aes(x1, response)) +
  geom_smooth(aes(col = factor(response_label))) +
  geom_point(aes(col = factor(response_label)))

ggplot(data_geo, aes(s1, s2)) +
  geom_point(aes(col = mgp.list.mean, size = mgp.list.mean)) +
  facet_wrap(~response_label)

vg <- data_geo %>%
  group_by(response_label) %>%
  nest() %>%
  mutate(variog = purrr::map(data, ~ gstat::variogram(mgp.list.mean ~ 1, ~ s1 + s2, . ,
                                                      cutoff = 3, width = 0.005))) %>%
  dplyr::select(-data) %>%
  unnest()

ggplot(vg, aes(dist, gamma)) +
  geom_point(aes(size = np)) +
  geom_smooth() +
  expand_limits(y = 0, x = 0) +
  scale_x_continuous(limits = c(0, 3)) +
  scale_y_continuous(limits = c(0, 1.3)) +
  stat_function(fun = function(x) 0.6 * (1-exp(- x/0.05)), col = 2, size = 1) +
  stat_function(fun = function(x) 0.6 * (1-exp(- x/0.1)), col = 3, size = 1) +
  facet_wrap(~response_label, ncol = 1)

vg <- data_geo %>%
  group_by(response_label) %>%
  nest() %>%
  mutate(variog = purrr::map(data, ~ gstat::variogram(response ~ 1, ~ s1 + s2, . ,
                                                      cutoff = 3, width = 0.005))) %>%
  dplyr::select(-data) %>%
  unnest()

ggplot(vg, aes(dist, gamma)) +
  geom_point(aes(size = np)) +
  geom_smooth() +
  expand_limits(y = 0, x = 0) +
  scale_x_continuous(limits = c(0, 3)) +
  scale_y_continuous(limits = c(0, 2)) +
  stat_function(fun = function(x) 0.4 + 0.6 * (1-exp(- x/0.05)), col = 2, size = 1) +
  stat_function(fun = function(x) 0.4 + 0.6 * (1-exp(- x/0.1)), col = 3, size = 1) +
  facet_wrap(~response_label, ncol = 1)

data_geo %>%
  dplyr::select(id, mre.factor.mean, response_label) %>%
  spread(response_label, mre.factor.mean) %>%
  dplyr::select(-id) %>%
  GGally::ggpairs(aes(fill = "any"))

data_geo_wide <- data_geo %>%
  dplyr::rename(ability = response, id_person = id) %>%
  gather(var, value, mre.factor.mean:ability) %>%
  mutate(var = paste0(var, response_label)) %>%
  select(-response_label) %>%
  spread(var, value)
```

### SIMULATE ITEM FACTOR DATA

```{r}
q <- 10
init_data <- purrr::map(1:q, ~ data_geo_wide) %>%
  purrr::reduce(rbind)

# n <- 300
difficulty <- matrix((1:q - 5)/10 * 2, nrow = 1)
discrimination1 <- seq(0.4, 1.5, length.out = q)
discrimination2 <- runif(q, 0, 2)
discrimination3 <- runif(q, 0, 2)
discrimination1[1] <- 1
discrimination1[c(3, 5, 8)] <- 0
discrimination2[1:2] <- c(0, 1)
discrimination2[c(4, 5, 10)] <- 0
# discrimination3[1:3] <- c(0, 0, 1)
# discrimination1 <- discrimination1 * 0.3
# discrimination2 <- discrimination2 * 0.3
cbind(discrimination1, discrimination2, discrimination3)

f <- list(
  prob ~ mfa(ones, beta = get("difficulty")) +
    mfe(ability1, beta = get("discrimination1")) +
    mfe(ability2, beta = get("discrimination2")),
  # + mfe(ability3, beta = get("discrimination3")),
  size ~ I(1)
  )

data_long <- sim_model(formula = f,
                        link_inv = list(pnorm, identity),
                        generator = rbinom,
                        responses = q,
                        n = n,
                        init_data = init_data
                        )

data_long <- dplyr::rename(data_long, subject = id,
                           item = response_label, y = response)

```


#### VISUALIZE ITEM FACTOR DATA

```{r}
explor <- data_long %>%
  group_by(subject) %>%
  summarize(endorse = mean(y),
            ability1 = unique(ability1),
            ability2 = unique(ability2),
            # ability3 = unique(ability3),
            x1 = unique(x1))
ggplot(explor, aes(ability1, endorse)) + geom_point(alpha = 0.5)
ggplot(explor, aes(ability2, endorse)) + geom_point(alpha = 0.5)
# ggplot(explor, aes(ability3, endorse)) + geom_point(alpha = 0.5)
# ggplot(explor, aes(x1, endorse)) + geom_point(alpha = 0.5)


```

### PREPARE DATA

```{r}

response <- data_long$y
coordinates <- dplyr::select(data_geo_wide, s1)
dist <- as.matrix(dist(coordinates))
# dist <- as.matrix(dist(dplyr::select(data_geo_wide, s1)[order(data_geo_wide$s1),]))
# dist <- dist[order(data_geo_wide$s1),]
n
q
m <- 2
iter <- 5 * 10 ^ 4
thin <- 5
# iter <- 5 * 10 ^ 3
cor.params <- c(0.04, 0.04)
sig.params <- c(0.6 ^ 0.5, 0.6 ^ 0.5)
fix.sigma <- 0.4^0.5
# sigma_prop <- matrix(c(0.138, -0.023, -0.023, 0.1), 2) * 2.38 ^ 2 / 2
sigma_prop <- 0.001 * diag(5)
disc_mat <- cbind(discrimination1, discrimination2)
L_a <- lower.tri(disc_mat, diag = TRUE) * 1
L_a[c(3,5,8), 1] <- 0
L_a[c(4,5,10), 2] <- 0
T_gp <- diag(m)
# diag(T_gp) <- 0
# T_gp[2,2] <- 0

```

### RUN

```{r}

data_test <- as_tibble(matrix(data_long$y, n, q)) %>%
  setNames(paste0("y", 1:q)) %>%
  mutate(s1 = data_geo_wide$s1, s2 = data_geo_wide$s2)

data_test_geo <- sf::st_as_sf(data_test, coords = c("s1", "s2"))

# iter <- 2000
# iter <- 10^5
# thin <- 50
iter <- 2 * 10^3
thin <- 1

system.time(
  samples_cifa <- spifa(
    responses = y1:y10, data = data_test, nfactors = 2,
    niter = iter, thin = thin, standardize = TRUE,
    constrains = list(A = L_a),
    adaptive = list(Sigma = NULL, Sigma_R = NULL, Sigma_gp_sd = NULL,
                    Sigma_gp_phi = NULL, scale = 1, C = 0.7, alpha = 0.8,
                    accep_prob = 0.234)
    )
  )

system.time(
  samples <- spifa(
    responses = y1:y10, data = data_test_geo, nfactors = 2,
    niter = iter, thin = thin, standardize = TRUE,
    constrains = list(A = L_a, W = T_gp, V_sd = 1), #sigmas[1:2]/2),
    adaptive = list(Sigma = NULL, Sigma_R = NULL, Sigma_gp_sd = NULL,
                    Sigma_gp_phi = NULL, scale = 1, C = 0.7, alpha = 0.8,
                    accep_prob = 0.234),
    sigmas_gp_opt = list(initial = 0.6, prior_mean = 0.6, prior_sd = 0.4),
    phi_gp_opt = list(initial = 0.1, prior_mean = 0.1, prior_sd = 0.4))
  )

save(list = c("samples", "samples_cifa"), file = "gp_samples.Rdata")

# load(file = "gp_samples.Rdata")

```


```{r}

iter <- iter / thin
thin2 <- 1

attr(samples, "model_info")[-c(1, 2, 3, 4)]
attr(samples_cifa, "model_info")[-c(1, 2, 3, 4)]

samples_tib <- as_tibble(samples, iter/2)
#summary(samples_tib)
samples_long <- gather(samples_tib)

as_tibble(samples, 0, thin2, "c") %>%
  gg_trace(alpha = 0.6)

as_tibble(samples_cifa, 0, thin2, "c") %>%
  gg_trace(alpha = 0.6)

as_tibble(samples, 0, thin2, "a") %>%
  gg_trace(alpha = 0.6)

as_tibble(samples_cifa, 0, thin2, "a") %>%
  gg_trace(alpha = 0.6)

as_tibble(samples, iter/2, thin2, "a") %>%
  gg_density(alpha = 0.5, ridges = TRUE, aes(fill = Parameters), scale = 4)

as_tibble(samples_cifa, iter/2, thin2, "a") %>%
  gg_density(alpha = 0.5, ridges = TRUE, aes(fill = Parameters), scale = 4)


as_tibble(samples, iter/2, thin2, "theta") %>%
  dplyr::select(1:100) %>%
  gg_density(alpha = 0.5, ridges = TRUE, aes(fill = Parameters), scale = 4)

as_tibble(samples_cifa, iter/2, thin2, "theta") %>%
  dplyr::select(1:100) %>%
  gg_density(alpha = 0.5, ridges = TRUE, aes(fill = Parameters), scale = 4)


as_tibble(samples, 0, thin2, "theta") %>%
  select(1:10) %>%
  gg_trace(alpha = 0.6)

as_tibble(samples_cifa, 0, thin2, "theta") %>%
  select(1:10) %>%
  gg_trace(alpha = 0.6)

as_tibble(samples, 0, thin2, "corr") %>%
  gg_trace(alpha = 0.6)

as_tibble(samples_cifa, 0, thin2, "corr") %>%
  gg_trace(alpha = 0.6)


as_tibble.spifa.list(samples, 0, thin2, "mgp_sd") %>%
  gg_trace(alpha = 0.6)

as_tibble.spifa.list(samples, iter/2, thin2, "mgp_sd") %>%
  gg_density(alpha = 0.5, ridges = FALSE, aes(fill = Parameters), scale = 4) +
  stat_function(fun = dlnorm, colour = "red",
                args = list(meanlog = log(0.6), sdlog = 0.4))

as_tibble.spifa.list(samples, 0, thin2, "mgp_phi") %>%
  gg_trace(alpha = 0.6)

as_tibble.spifa.list(samples, iter/2, thin2, "mgp_phi") %>%
  gg_density(alpha = 0.5, ridges = FALSE, aes(fill = Parameters), scale = 4) +
  stat_function(fun = dlnorm, colour = "red",
                args = list(meanlog = log(0.1), sdlog = 0.4))

as_tibble.spifa.list(samples, 0, thin2, "a") %>%
  gg_density2d(`Discrimination 1`, `Discrimination 2`, each = 10,
               keys = c("Item ", "Discrimination "),
               highlight = c(discrimination1, discrimination2))

as_tibble.spifa.list(samples, 0, thin2, "a") %>%
  gg_scatter(`Discrimination 1`, `Discrimination 2`, each = 10,
               keys = c("Item ", "Discrimination "),
               highlight = c(discrimination1, discrimination2))

as_tibble.spifa.list(samples, iter/ 2, select = "a") %>%
  summary() %>%
  mutate(param = c(discrimination1, discrimination2)) %>%
  gg_errorbarh() +
  geom_point(aes(param, Parameters), col = 3)

as_tibble.spifa.list(samples, iter/2, select = "c") %>%
  summary() %>%
  mutate(param = as.numeric(difficulty)) %>%
  gg_errorbarh() +
  geom_point(aes(param, Parameters), col = 3)

as_tibble.spifa.list(samples, iter/2, select = "theta") %>%
  dplyr::select(1:300) %>%
  summary() %>%
  mutate(param = data_geo$response[1:300]) %>%
  gg_errorbarh(sorted = TRUE) +
  geom_point(aes(x = param), col = 3)

as_tibble.spifa.list(samples, iter/2, select = "theta") %>%
  dplyr::select(301:600) %>%
  summary() %>%
  mutate(param = data_geo$response[301:600]) %>%
  gg_errorbarh(sorted = TRUE) +
  geom_point(aes(x = param), col = 3)

ability1_pred <- as_tibble.spifa.list(samples, iter/2, select = "theta") %>%
  dplyr::select(1:300) %>%
  summary() %>%
  mutate(param = data_geo$response[1:300],
         s1 = data_geo$s1[1:300],
         s2 = data_geo$s2[1:300],
         estim = `50%`)
ability1_pred %>%
    ggplot(aes(s1, `50%`)) +
    geom_line() +
    geom_line(aes(s1, param, col = "real"))

vg <- gstat::variogram(estim ~ 1, ~ s1 + s2, ability1_pred, cutoff = 3, width = 0.01)
ggplot(vg, aes(dist, gamma)) +
  geom_point(aes(size = np)) +
  geom_smooth() +
  expand_limits(y = 0, x = 0) +
  scale_x_continuous(limits = c(0, 3)) +
  stat_function(fun = function(x) 0.6 * (1-exp(- x/0.05)), col = 2, size = 1)

ability1_pred <- as_tibble.spifa.list(samples, iter/2, select = "theta") %>%
  dplyr::select(1:300)
ability1_pred <- ability1_pred[nrow(ability1_pred),]
ability1_pred <- ability1_pred %>%
  spmirt:::summary.spifa() %>%
  mutate(param = data_geo$response[1:300],
         s1 = data_geo$s1[1:300],
         s2 = data_geo$s2[1:300],
         estim = `50%`)
ability1_pred %>%
    ggplot(aes(s1, `50%`)) +
    geom_line() +
    geom_line(aes(s1, param, col = "real"))

vg <- gstat::variogram(estim ~ 1, ~ s1 + s2, ability1_pred, cutoff = 3, width = 0.01)
ggplot(vg, aes(dist, gamma)) +
  geom_point(aes(size = np)) +
  geom_smooth() +
  expand_limits(y = 0, x = 0) +
  scale_x_continuous(limits = c(0, 3))

ability2_pred <- as_tibble.spifa.list(samples, iter/2, select = "theta") %>%
  dplyr::select(301:600) %>%
  spmirt:::summary.spifa() %>%
  mutate(param = data_geo$response[301:600],
         s1 = data_geo$s1[301:600],
         s2 = data_geo$s2[301:600],
         estim = `50%`)
ability2_pred %>%
  ggplot(aes(s1, `50%`)) +
  geom_line() +
  geom_line(aes(s1, param, col = "real"))

vg <- gstat::variogram(estim ~ 1, ~ s1 + s2, ability2_pred, cutoff = 3, width = 0.005)
ggplot(vg, aes(dist, gamma)) +
  geom_point(aes(size = np)) +
  geom_smooth() +
  expand_limits(y = 0, x = 0) +
  scale_x_continuous(limits = c(0, 2))

ability2_pred <- as_tibble.spifa.list(samples, iter/2, select = "theta") %>%
  dplyr::select(301:600)
# ability2_pred <- ability2_pred[nrow(ability2_pred),]
ability2_pred <- ability2_pred %>%
  spmirt:::summary.spifa() %>%
  mutate(param = data_geo$response[301:600],
         s1 = data_geo$s1[301:600],
         s2 = data_geo$s2[301:600],
         estim = `50%`)
ability2_pred %>%
  ggplot(aes(s1, `50%`)) +
  geom_line() +
  geom_line(aes(s1, param, col = "real"))
vg <- gstat::variogram(estim ~ 1, ~ s1 + s2, ability2_pred, cutoff = 3, width = 0.005)
ggplot(vg, aes(dist, gamma)) +
  geom_point(aes(size = np)) +
  geom_smooth() +
  expand_limits(y = 0, x = 0) +
  scale_x_continuous(limits = c(0, 2)) +
  scale_y_continuous(limits = c(0, 2))

```

### Model selection

```{r}
dic(as_tibble(samples_cifa, burnin = iter/2, thin = 1))
dic(as_tibble(samples, burnin = iter/2, thin = 1))
```

### Prediction

```{r}

library(sf)
# new prediction points
margins <- st_bbox(data_test_geo)
tol <- 0.01
ngrid <- 30
long_seq <- seq(margins[1] - tol, margins[3] + tol, length = ngrid)
lat_seq <- seq(margins[2] - tol, margins[4] + tol, length = ngrid)
newcoords <- expand.grid(gps_longitude = long_seq, gps_latitude = lat_seq) %>%
  sf::st_as_sf(coords = c("gps_longitude", "gps_latitude"), crs = st_crs(data_test_geo))

ggplot(newcoords) +
  geom_sf()

newcoords_mat <- sf::st_coordinates(newcoords)

system.time(
bla <- as_tibble(samples, burnin = iter/2, thin = 20) %>%
  predict(newcoords = newcoords_mat)
)


# SUMMARIZE PREDICTION AND JOIN WITH ORIGINAL DATASET --------------------------

theta_excees <- as_tibble.spifa.list(bla) %>%
  purrr::map_dbl( ~ sum(. > 0) / length(.)) %>%
  as.numeric()

theta_quant <- as_tibble.spifa.list(bla) %>%
  summary() %>%
  mutate(exceed_prob = theta_excees) %>%
  tidyr::extract(Parameters, c("subject", "factor"), "Theta\\[([0-9]+),([0-9]+)\\]",
                 convert = TRUE) %>%
  tidyr::gather(quantile, value, `2.5%`:`exceed_prob`, factor_key = TRUE) %>%
  dplyr::mutate(
    factor_quantile = factor(paste0("F", factor, "-", quantile),
                      paste0("F", rep(1:m, each = 6), "-", rep(levels(quantile), m)))
    ) %>%
  dplyr::select(-factor, -quantile) %>%
  tidyr::spread(factor_quantile, value) %>%
  dplyr::arrange(subject)

plop <- dplyr::bind_cols(newcoords, theta_quant)

# scaling
mean_scale <- plop %>%
  select(`F1.10.`, `F1.50.`, `F1.90.`) %>%
  st_set_geometry(NULL) %>%
  colMeans() %>%
  mean()
scale_center_trans = function() {
  trans_new("scale_center", function(x) x - mean_scale, function(x) x + mean_scale)}

# ggplot
plop %>%
  ggplot() +
  geom_sf(aes(col = `F1-50%`))


# newcoords <- matrix(seq(0, 2, 0.02))
#
# system.time(
# bla <- predict(samples, newcoords = newcoords, burnin = iter/2, thin = 10)
# )
#
# as_tibble.spifa.list(bla) %>%
#   dplyr::select(1:30) %>%
#   gg_trace()
#
# as_tibble.spifa.list(bla) %>%
#   dplyr::select(1:80) %>%
#   gg_density(ridges = TRUE)
#
# as_tibble.spifa.list(bla) %>%
#   spmirt:::summary.spifa() %>%
#   mutate(
#   s1 = rep(as.numeric(newcoords), 2),
#   response_label = rep(1:m, each = nrow(newcoords))
#   ) %>%
#   ggplot() +
#   geom_line(aes(s1, `50%`, col = "x1")) +
#   geom_ribbon(aes(x = s1, ymin = `2.5%`, ymax = `97.5%`), alpha = 1/4) +
#   geom_line(aes(s1, mgp.list.mean), subset(data_geo, response_label < 3)) +
#   facet_wrap(~response_label)
```



