---
title: "Spatial Item Factor Analysis"
author: "Erick A. Chacon-Montalvan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatial Item Factor Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")
knitr::opts_chunk$set(fig.width = 7, fig.height = 4)
knitr::opts_chunk$set(comment = "#>")
options(width = 100)
```

## Introduction

In this vignette, we show how to use the **spifa** package to fit a spatial
multidimensional 2 parameter logistic model, as used in item response
theory. Let $Y_{ij}$ be the binary response for item $j$ in individual $i$.
The model can be defined using an auxiliary variable $Z_{ij}$ such that

\begin{align}
  {Y}_{ij}  & =
  \left\lbrace
  \begin{array}[2]{cc}
    1 & \text{if} ~ {Z}_{ij} > 0\\
    0 & \text{otherwise}
  \end{array}
  \right.\\
  {Z}_{ij} & = c_j + a_j^\intercal\theta_i + \epsilon_{ij},
  ~~ \epsilon_{ij} \sim {N}(0, 1)
\end{align}

where the latent abilities $\theta_i$ are, in the spatial case, explained by
a multivariate Gaussian process rather than assumed independent across
individuals.

## Load required packages and data

```{r}
library(dplyr)
library(ggplot2)
library(spifa)
library(sf)

data(ipixuna)
```

`ipixuna_wide` has one row per (simulated) household, with ten binary food
insecurity items, a covariate `x1`, and a spatial coordinate `coords`:

```{r}
ipixuna_wide %>%
  dplyr::select(-coords) %>%
  head(10) %>%
  knitr::kable(digits = 2)
```

The true generating parameters used to simulate the data (including the
discrimination structure) are stored as an attribute, which is convenient
both for setting sensible restrictions and for checking recovery below:

```{r}
parameters <- attr(ipixuna_wide, "parameters")
str(parameters)
```

## Fit a SPIFA model for the Ipixuna data

We restrict the discrimination matrix to the structure used to simulate the
data, and let the two latent factors follow independent Gaussian processes
(`W = diag(2)`) over the household coordinates. `niter` is kept small here
so the vignette builds quickly; in practice use a few thousand iterations
and check convergence (see `?gg_trace`).

```{r}
L_a <- (parameters$discrimination != 0) * 1

ipixuna_sf <- sf::st_as_sf(ipixuna_wide)
ipixuna_sf$items <- as.matrix(dplyr::select(ipixuna_wide, `Item 1`:`Item 10`))

system.time(
  samples <- spifa(
    items ~ x1, data = ipixuna_sf, nfactors = 2,
    niter = 200, thin = 1, standardize = FALSE,
    constraints = list(discrimination = L_a, mgp = diag(2), resid_sd = rep(0.4^0.5, 2)),
    priors = list(
      mgp_sd = list(initial = 0.6, mean = 0.6, sd = 0.4),
      mgp_range = list(initial = 200, mean = 200, sd = 0.4)))
  )

attr(samples, "model_info")$model_type
```

## Summarise samples

```{r}
summary(samples, burnin = 100, select = "c")
```

## Visualize results

Traceplots of the easiness (`c`) and discrimination (`a`) parameters:

```{r, fig.height = 6}
as_tibble(samples, select = "c") %>%
  gg_trace(alpha = 0.6)

as_tibble(samples, select = "a") %>%
  gg_trace(alpha = 0.6)
```

Traceplots of the residual correlation and of the Gaussian process
hyperparameters (standard deviation and range):

```{r, fig.height = 6}
as_tibble(samples, select = "corr") %>%
  gg_trace(alpha = 0.6)

as_tibble(samples, select = "mgp_sd") %>%
  gg_trace(alpha = 0.6)

as_tibble(samples, select = "mgp_phi") %>%
  gg_trace(alpha = 0.6)
```

Credible intervals for the discrimination parameters, against the true
simulated values:

```{r, fig.height = 5}
summary(samples, burnin = 100, select = "a") %>%
  mutate(param = as.numeric(parameters$discrimination)) %>%
  gg_errorbarh(sorted = FALSE) +
  geom_point(aes(x = param), col = 3)
```
