---
title: "Spatial Item Factor Analysis"
author: "Erick A. Chacon-Montalvan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatial Item Factor Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")
knitr::opts_chunk$set(fig.width = 7, fig.height = 4)
knitr::opts_chunk$set(comment = "#>")
options(width = 100)
```

## Introduction

In this vignette, we show how to use the **spifa** package to fit a
multidimensional 2 parameter logistic model used in item response theory. Let
$Y_{ij}$ be the response for item
$j$ in the individual $i$. The model can be defined by using an auxiliary variable
$Z_{ij}$ such as
\begin{align}
  {Y}_{ij}  & =
  \left\lbrace
  \begin{array}[2]{cc}
    1 & \text{if} ~ {Z}_{ij} > 0\\
    0 & \text{otherwise}
  \end{array}
  \right.\\
  {Z}_{ij} & = c_j + a_j^\intercal\theta_i + \epsilon_{ij},
  ~~ \epsilon_{ij} \sim {N}(0, 1)\\
  {\theta}_i & \sim {N}({0}, {I}_m)\\
  c_j & \sim {N}(0, \sigma_c^2)\\
  {a}_j & \sim {N}(0, \sigma_a^2{I}_m)\\
\end{align}

## Load required packages and data

```{r}
library(dplyr)
library(ggplot2)
library(spifa)

data(ipixuna)
# ipixuna_wide <- ipixuna_wide %>%
#   sf::st_sf()
#   dplyr::select(-coords)
# ipixuna_wide[c("lon", "lat")] <- list(rnorm(200), rnorm(200))
```

```{r}
knitr::kable(head(ipixuna, 10))
```

```{r}
knitr::kable(head(ipixuna_wide, 10))
```

## Fit a spifa model for Ipixuna data

```{r}
# data(ipixuna)
# ipixuna_wide <- sf::st_sf(ipixuna_wide)


# parameters used for simulation
parameters <- attr(ipixuna, "parameters")

# restrictions
L_a <- (parameters$discrimination != 0) * 1
iter <- 1000

# system.time(
#   samples <- spifa(
#     responses = `Item 1`:`Item 10`, data = ipixuna_wide, nfactors = 2,
#     niter = iter, thin = 1, standardize = FALSE,
#     constrains = list(A = L_a, W = diag(2), V_sd = 0.3 ^ 0.5),
#     adaptive = list(Sigma = NULL, Sigma_R = NULL, Sigma_gp_sd = NULL,
#                     Sigma_gp_phi = NULL, scale = 1, C = 0.7, alpha = 0.8,
#                     accep_prob = 0.234),
#     sigmas_gp_opt = list(initial = 0.6, prior_mean = 0.6, prior_sd = 0.4),
#     phi_gp_opt = list(initial = 100, prior_mean = 100, prior_sd = 4))
#   )

system.time(
  samples <- spifa(
    responses = `Item 1`:`Item 10`, pred_formula = ~ x1, coords = coords,
    data = ipixuna_wide, nfactors = 2,
    niter = 100, thin = 1, standardize = FALSE,
    constrains = list(A = L_a, W = diag(2), V_sd = 0.4 ^ 0.5),
    adaptive = list(Sigma = NULL, Sigma_R = NULL, Sigma_gp_sd = NULL,
                    Sigma_gp_phi = NULL, scale = 1, C = 0.7, alpha = 0.8,
                    accep_prob = 0.234),
    sigmas_gp_opt = list(initial = 0.6, prior_mean = 0.6, prior_sd = 0.4),
    phi_gp_opt = list(initial = 200, prior_mean = 200, prior_sd = 0.4),
    execute = TRUE)
  )


attr(samples, "model_info")

# hist(exp(rnorm(50000, log(0.6), 0.4)), 50)

# unclass(attr(samples, "model_info")$distance[1:5, 1:5])

```

## Visualize results

```{r}
# library(purrr)
# samples_tib <- as_tibble(samples)
#
# # summary(samples_tib)
# # samples_long <- gather(samples_tib)
# thin2 <- 1
#
# as_tibble(samples, 0, thin2, "c") %>%
#   gg_trace(alpha = 0.6)
#
# as_tibble(samples, 0, thin2, "a") %>%
#   gg_trace(alpha = 0.6)
#
# as_tibble(samples, 0, thin2, "corr") %>%
#   gg_trace(alpha = 0.6)
#
# as_tibble.spifa.list(samples, 0, thin2, "mgp_sd") %>%
#   gg_trace(alpha = 0.6)
#
# as_tibble.spifa.list(samples, 0, thin2, "mgp_phi") %>%
#   gg_trace(alpha = 0.6)
#
```

