---
title: "Spatial Item Factor Analysis"
author: "Erick A. Chacon-Montalvan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatial Item Factor Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")
knitr::opts_chunk$set(fig.width = 7, fig.height = 4)
knitr::opts_chunk$set(comment = "#>")
options(width = 100)
```

## Introduction

In this vignette, we show how to use the **spifa** package to fit a spatial
multidimensional 2 parameter logistic model, as used in item response
theory. Let $Y_{ij}$ be the binary response for item $j$ in individual $i$.
The model can be defined using an auxiliary variable $Z_{ij}$ such that

\begin{align}
  {Y}_{ij}  & =
  \left\lbrace
  \begin{array}[2]{cc}
    1 & \text{if} ~ {Z}_{ij} > 0\\
    0 & \text{otherwise}
  \end{array}
  \right.\\
  {Z}_{ij} & = c_j + a_j^\intercal\theta_i + \epsilon_{ij},
  ~~ \epsilon_{ij} \sim {N}(0, 1)
\end{align}

where the vector of latent abilities $\theta_i$ is modelled, in the spatial
case, as

$$
\theta_i = X_i^\intercal B + w(s_i) + v_i,
$$

the sum of a predictor effect ($X_i^\intercal B$, with $X_i$ the row of
predictors for individual $i$ and $B$ the matrix of effects), a
multivariate Gaussian process capturing spatial dependence ($w(s_i)$, at
location $s_i$), and a multivariate non-spatial term ($v_i$), rather than
assumed independent across individuals.

## Load required packages and data

```{r}
library(dplyr)
library(ggplot2)
library(spifa)
library(sf)

data(ipixuna)
```

`ipixuna` has one row per (simulated) household, with a ten-item binary
`items` response matrix, a covariate `wealth`, and a spatial coordinate
`geometry`:

```{r}
ipixuna |>
  dplyr::select(-geometry, -items) |>
  head(10) |>
  knitr::kable(digits = 2)
```

The true generating parameters used to simulate the data (including the
discrimination structure) are stored as an attribute, which is convenient
both for setting sensible restrictions and for checking recovery below:

```{r}
parameters <- attr(ipixuna, "parameters")
str(parameters)
```

## Fit a SPIFA model for the Ipixuna data

We restrict the discrimination matrix to the structure used to simulate the
data, and let the three latent factors follow independent Gaussian
processes (`W = diag(3)`) over the household coordinates. `niter` is kept
small here so the vignette builds quickly; in practice use a few thousand
iterations and check convergence (see `?plot_trace`).

```{r}
L_a <- (parameters$discrimination != 0) * 1
nfactors <- ncol(parameters$discrimination)

system.time(
  samples <- spifa(
    items ~ wealth, data = ipixuna, nfactors = nfactors,
    niter = 1000, thin = 1, standardize = FALSE,
    constraints = list(discrimination = L_a, loading = diag(nfactors),
      sd = parameters$resid_params$sd),
    priors = list(
      loading = list(initial = 0.6, mean = 0.6, sd = 0.4),
      range = list(initial = 200, mean = 200, sd = 0.4)))
  )
samples

attr(samples, "fit_args")$model_type
```

## Summarise samples

```{r}
summary(samples, burnin = 100, select = "c")
```

## Visualize results

Traceplots of the easiness (`c`) and discrimination (`a`) parameters:

```{r, fig.height = 6}
plot_trace(samples, select = "c")

plot_trace(samples, select = "A")
```

Traceplots of the residual correlation and of the Gaussian process
hyperparameters (loading matrix `T` and range `phi`):

```{r, fig.height = 6}
plot_trace(samples, select = "Corr")

plot_trace(samples, select = "T")

plot_trace(samples, select = "phi")
```

Credible intervals for the discrimination parameters, against the true
simulated values:

```{r, fig.height = 5}
plot_interval(samples, select = "A", burnin = 100,
              reference = parameters$discrimination)
```

## Predict and map the latent factors

The reason **spifa** models the latent factors as spatial processes is to
predict them at new, unobserved locations -- this is what makes it possible
to map a construct like food insecurity across a whole area from survey
data collected at a limited number of households. `predict()` takes an
`sf`/`sfc` object of new locations (with any predictor columns the model
needs, `wealth` here) and returns posterior predictive draws of the latent
factors at those locations:

```{r, fig.height = 5}
newdata <- st_sf(wealth = 0, geometry = st_make_grid(ipixuna, n = c(15, 15),
  what = "centers"))
pred <- predict(samples, newdata = newdata, burnin = 100)

newdata$theta1 <- posterior::summarise_draws(pred, mean)$mean[
  grepl(",1\\]$", posterior::variables(pred))]

ggplot(newdata) +
  geom_sf(aes(colour = theta1)) +
  scale_colour_viridis_c() +
  labs(colour = "Factor 1\n(predicted mean)")
```

## Compare models with DIC

`dic()` computes the Deviance Information Criterion, useful for comparing
candidate models fitted to the same data -- e.g. to choose the number of
factors, or to compare restriction structures. Here we compare the
3-factor structure used above against a 2-factor alternative (both without
the spatial/predictor structure, so the comparison is fast):

```{r}
nitems <- ncol(ipixuna$items)
A3 <- matrix(1, nitems, 3)
A3[c(1, 3, 5), 1] <- 0
A3[c(2, 6, 9), 2] <- 0
A3[c(4, 7, 10), 3] <- 0

A2 <- matrix(1, nitems, 2)
A2[c(1, 3, 5, 9), 1] <- 0
A2[c(2, 4, 6, 7, 10), 2] <- 0

fit3 <- spifa(items ~ 1, data = ipixuna, nfactors = 3, ngp = 0, niter = 300,
  constraints = list(discrimination = A3))
fit2 <- spifa(items ~ 1, data = ipixuna, nfactors = 2, ngp = 0, niter = 300,
  constraints = list(discrimination = A2))

dplyr::bind_rows(
  "3 factors" = dic(fit3, burnin = 100),
  "2 factors" = dic(fit2, burnin = 100),
  .id = "model")
```

The 3-factor model has the lower DIC, correctly favouring the structure
used to simulate the data.
