---
title: "Preparing Your Data"
author: "Erick A. Chacon-Montalvan"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preparing Your Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center")
knitr::opts_chunk$set(comment = "#>")
options(width = 100)
```

## Introduction

`spifa()` expects data in a specific shape: one row per respondent, and the
binary item responses stored as a single *matrix-valued column* (`nobs x
nitems`) rather than one column per item. This is the same convention base
R uses for multivariate `lm()`. Survey data rarely arrives in that shape --
this vignette builds a small questionnaire dataset from scratch, the way
it typically comes out of a survey tool (one column per item, `"Yes"`/
`"No"` responses, a household covariate, GPS coordinates), and walks
through getting it into the shape `spifa()` needs.

```{r}
library(dplyr)
library(sf)
library(spifa)
```

## Simulating a raw questionnaire export

Five binary items, a household `wealth` covariate, and GPS coordinates,
for 20 households -- exactly what a survey export typically looks like:

```{r}
set.seed(1)
nobs <- 20
nitems <- 5

raw <- data.frame(
  id = seq_len(nobs),
  wealth = rnorm(nobs),
  lon = runif(nobs, -71.70, -71.68),
  lat = runif(nobs, -7.06, -7.04))

for (j in seq_len(nitems)) {
  raw[[paste0("item", j)]] <- factor(ifelse(rbinom(nobs, 1, 0.5) == 1, "Yes", "No"))
}

head(raw, 3)
```

## From one column per item to a matrix column

Recode each item to binary `0`/`1` (`spifa()` also accepts `TRUE`/`FALSE`,
coerced the same way), then assemble the item columns into a single matrix
column with `as.matrix()` + ordinary `$<-` assignment (or `I()`), which
preserves it as a matrix rather than flattening it back into separate
columns:

```{r}
data <- raw |>
  mutate(across(starts_with("item"), ~ as.integer(.x == "Yes")))

data$items <- as.matrix(select(data, starts_with("item")))
data <- select(data, id, wealth, lon, lat, items)

str(data)
```

Passing the original one-column-per-item `raw` data (before this step) to
`spifa()` fails clearly rather than silently doing the wrong thing:

```{r, error = TRUE}
spifa(item1 ~ 1, data = raw, nfactors = 2, niter = 5)
```

## Missing item responses

Individual missing item responses (`NA`) don't need to be dropped or
imputed -- `spifa()` handles them natively, treating a missing response as
an unobserved auxiliary variable to be sampled along with everything else,
rather than requiring complete rows:

```{r}
data_na <- data
data_na$items[1, 1] <- NA

fit <- spifa(items ~ 1, data = data_na, nfactors = 2, ngp = 0, niter = 5)
attr(fit, "fit_args")$nobs  # respondent with the NA item is still included
```

## Missing predictor values

Missing *predictor* values behave differently, and are worth calling out:
a `NA` in a predictor column drops that respondent from the whole fit, the
same way `lm()` and friends do -- unlike item responses, there's no
partial information to keep:

```{r}
data_na <- data
data_na$wealth[1] <- NA

fit <- spifa(items ~ wealth, data = data_na, nfactors = 2, ngp = 0, niter = 5)
attr(fit, "fit_args")$nobs  # one respondent dropped
```

## Adding spatial coordinates

Spatial models need `data` to be an `sf` object. If your coordinates are
plain numeric columns (e.g. from a GPS device), as here, convert with
`sf::st_as_sf()`:

```{r}
data_sf <- st_as_sf(data, coords = c("lon", "lat"), crs = 4326)
```

`spifa()` then adds the spatial Gaussian process automatically (see
`vignette("spifa")`):

```{r}
attr(spifa(items ~ 1, data = data_sf, nfactors = 2, niter = 5),
  "fit_args")$model_type
```

## Putting it together

```{r}
fit <- spifa(items ~ wealth, data = data_sf, nfactors = 2, niter = 5)
fit
```
