---
title: "Model types"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model types}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center", out.width = "100%")
knitr::opts_chunk$set(comment = "#>")
options(width = 100)
```

`spifa()` can fit five model types that are determined automatically from `formula` and
the class of `data` (see `?spifa`):

- Non-spatial:
  - `eifa`: Exploratory IFA
  - `cifa`: Confirmatory IFA
  - `cifa_pred`: CIFA with predictors
- Spatial:
  - `spifa`: Spatial IFA
  - `spifa_pred`: SPIFA with predictors

The model type is determined based on the following:

1) Exploratory structure (`eifa`) is used when no `constraints` are provided.
2) Models with predictors (`cifa_pred`/`spifa_pred`) are activated if terms are included
on the right-hand side of `formula`.
3) Spatial models (`spifa`/`spifa_pred`) are only activated when the `data` is
geo-referenced as an `sf` object, otherwise the non-spatial models are activated.
4) Setting `ngp = 0` forces the model to be non-spatial (`eifa`/`cifa`/`cifa_pred`).

We will showcase the five model types using the `ipixuna` data.

```{r}
library(spifa)

data(ipixuna)
nitems <- ncol(ipixuna$items)
```

## Exploratory item factor analysis (EIFA)

The simplest case is the `eifa` model, the user does not need to provide `constraints`
neither a spatial `data`. The constraints will be internally defined as a lower-triangular matrix; and in case `data` is an `sf` object, the spatial functionality can
be dropped using `ngp = 0`:

```{r}
set.seed(123)
samples <- spifa(items ~ 1, data = ipixuna, nfactors = 2, ngp = 0, niter = 500)
samples
```

Notice how the constraints have been defined internally:

```{r}
attr(samples, "fit_args")$constrain_L
```

EIFA also never estimates the correlation between the latent factors, so only the
easiness (c) and discrimination (A) parameters are estimated.

## Confirmatory item factor analysis (CIFA)

Supplying your own discrimination `constraints` leads to a `cifa` model:

```{r}
nfactors <- 3
A <- matrix(1, nitems, nfactors)
A[c(4, 8), 1] <- 0
A[c(4, 5, 6, 7, 8, 10), 2] <- 0
A[c(5, 6), 3] <- 0

samples <- spifa(
  items ~ 1, data = ipixuna, nfactors = nfactors, ngp = 0, niter = 500,
  constraints = list(discrimination = A)
)
samples
```

Unlike `eifa`, the `cifa` model estimates the correlation (Corr) between the latent factors.

## CIFA model with predictors

Adding predictor terms to the right-hand side of `formula` generates a `cifa_pred`, which models the latent factors in terms of the predictor terms:

```{r}
samples <- spifa(
  items ~ poly(wealth, 2), data = ipixuna, nfactors = nfactors, ngp = 0, niter = 500,
  constraints = list(discrimination = A)
)
samples
```

## Spatial item factor analysis (SPIFA)

If the `data` is an `sf` object and no predictor terms are included in `formula`, the
model type is `spifa` automatically. By default, it includes a spatial term for each
factor `ngp = nfactors`; but custom number of `ngp` can be provided with a constraints
`loading` matrix (T) of size `nfactors` x `ngp`.

```{r}
T_loading <- matrix(c(1, 1, 0, 0, 0, 1), nfactors, 2)
T_loading
```

This structure means that 1 GP is shared between factor 1 and 2, and another
GP is introduced for factor 3:

```{r}
samples <- spifa(
  items ~ 1, data = ipixuna, nfactors = nfactors, ngp = 2, niter = 500,
  constraints = list(discrimination = A, loading = T_loading)
)
samples
```

Notice that the `loading` parameter is also available in the samples.

## SPIFA model with predictors

A `spifa_pred` model is simply defined by adding predictor terms to `formula`.

```{r}
samples <- spifa(
  items ~ wealth, data = ipixuna, nfactors = nfactors, niter = 500,
  constraints = list(discrimination = A)
)
samples
```

Notice that the predictor effects `B` is now included in the samples.
