---
title: "Advanced features"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced features}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center", out.width = "100%")
knitr::opts_chunk$set(comment = "#>")
options(width = 100)
```

This vignette goes deeper into two things `vignette("spifa")` and
`vignette("vg02-model-types")` only touch briefly:

- how to control `constraints`/`priors` for every parameter group,
- how to continue sampling with `update()` when a fit hasn't converged yet, and
- how to compare models using `dic()`.

```{r}
library(spifa)

data(ipixuna)
nfactors <- 3
nitems <- ncol(ipixuna$items)
```

## Restrictions and priors

`spifa()`'s parameter glossary (see `?spifa`) maps six configurable parameter names --
`easiness` (c), `discrimination` (A), `effect` (B), `corr` (Corr), `loading` (T), `range`
(phi). Two `spifa()` arguments use these names:

- `constraints`: fixes cells of a parameter to a value instead of estimating them. Only
  `discrimination`, `loading`, and `sd` support this (the model needs some fixed structure
  for identifiability).
- `priors`: sets the `initial` value and prior hyperparameters (`mean`/`sd`,
  or `eta` for `corr`) for a parameter that is estimated. Every one of the six names
  accepts this.

### `constraints`

- `discrimination` (nitems x nfactors matrix): which items inform which
  factors,
- `loading` (nfactors x ngp matrix): which Gaussian processes feed into
  which factors, and
- `sd`: fixed values for the latent factors' residual standard deviation (length
  `nfactors`). Unlike `discrimination`/`loading`, this is not estimated, it is always
  fixed.

Define `discrimination` structure:

```{r}
A <- matrix(1, nitems, nfactors)
A[c(4, 8), 1] <- 0
A[c(4, 5, 6, 7, 8, 10), 2] <- 0
A[c(5, 6), 3] <- 0
A
```

Define `loading` structure, factors 1 and 2 share one Gaussian process, factor 3 gets its
own:

```{r}
ngp <- 2
T_loading <- matrix(c(1, 1, 0, 0, 0, 1), nfactors, ngp)
T_loading
```

Fix the residual standard deviation of each factor to 0.2:

```{r}
sd_fixed <- rep(0.2, nfactors)
sd_fixed
```

### `priors`

Every element accepts `initial` (starting value for the sampler) and the prior
hyperparameters `mean`/`sd`. The parameter `corr` is the exception: it takes `initial` and
`eta` (the LKJ shape parameter) instead.

By default, `spifa()` starts every parameter at a generic default and lets the data speak.
If you have domain knowledge, setting `priors` closer to a plausible value can speed up
convergence substantially. Here we set a plausible prior for every parameter group:

- item easiness roughly increasing across items,
- a couple of items with a strong sign-fixing discrimination prior (as in `vignette("spifa")`),
- a wealth effect that differs in sign across factors,
- moderate loadings,
- factor correlations nudged away from independence, and
- a range parameter around `150` as in `vignette("spifa")`:

```{r}
# easiness hyperparameters
c_mean <- seq(-0.7, 0.9, length.out = nitems)
c_sd <- rep(0.3, nitems)

# discrimination hyperparameters
A_mean <- matrix(0, nitems, nfactors)
A_mean[c(5, 6), 1] <- 1
A_mean[c(4, 8), 3] <- -1
A_sd <- matrix(1, nitems, nfactors)
A_sd[A_mean != 0] <- 0.45

# predictor effects hyperparameters
B_mean <- matrix(c(-0.5, 0.5, 0.2), nrow = 1)
B_sd <- matrix(0.5, nrow = 1, ncol = nfactors)

# correlation initial value
corr_initial <- diag(nfactors)

# range parameter hyperparameters
phi_mean <- 150
phi_sd <- 0.4
```

### Fit with constraints and priors

In the following code, we explicitely defined the `constraints`, and `priors` for all
possible parameters:

```{r}
set.seed(1)
fit_informed <- spifa(
  items ~ wealth, data = ipixuna, nfactors = nfactors, ngp = 3,
  burnin = 200, niter = 200, thin = 2,
  constraints = list(discrimination = A, loading = diag(nfactors), sd = sd_fixed),
  priors = list(
    easiness = list(initial = c_mean, mean = c_mean, sd = c_sd),
    discrimination = list(initial = A_mean, mean = A_mean, sd = A_sd),
    effect = list(initial = B_mean, mean = B_mean, sd = B_sd),
    corr = list(initial = corr_initial, eta = 2),
    loading = list(initial = 0.6, mean = 0.6, sd = 0.3),
    range = list(initial = phi_mean, mean = phi_mean, sd = phi_sd)))
fit_informed
```

## Continue sampling

Fits don't always converge on the first try. If you take a look at `rhat` in the previous
section, you will notice that some values are much higher that `1.05` for example. To
continue sampling from the current state you can use `update()`:

```{r, warning = FALSE}
fit1 <- update(fit_informed, niter = 1000)
fit1
```

`rhat` has moved closer to 1 with the extra iterations. In practice, you have to run many
more iterations.

## Compare models with DIC

`dic()` computes the Deviance Information Criterion for a fitted model, useful for
comparing candidate models fitted to the same data. You should compare them once
convergence is ensured. For example, let's compare with a `spifa` with 2 latent factors
and default `constraints` and `priors`:

```{r, warning = FALSE}
fit2 <- spifa(items ~ wealth, data = ipixuna, nfactors = 2, niter = 1000, burnin = 1000)
fit2
```

Computing `dic()` for each model:

```{r}
# M1: 3 factors
dic(fit1, burnin = 100, thin = 2)
```

```{r}
# M2: 2 factors
dic(fit2, burnin = 100, thin = 2)
```

Assuming both models have converged, the 3-factor model has the lower DIC, favouring the
discrimination structure used above over an 2-factor alternative.
