---
title: "Integrated workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Integrated workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center", out.width = "100%")
knitr::opts_chunk$set(fig.width = 7, fig.height = 4)
knitr::opts_chunk$set(comment = "#>")
options(width = 100)

can_run <- requireNamespace("mirt", quietly = TRUE) &&
  requireNamespace("gstat", quietly = TRUE)
knitr::opts_chunk$set(eval = can_run)
```

```{r, eval = !can_run, echo = FALSE, results = "asis"}
cat("This vignette needs the mirt and gstat packages, which are not installed.")
```

`vignette("spifa")` fits the full spatial model directly, using a discrimination
restriction and a spatial range prior already decided. In practice both could be
informed by previous knowledge or a preliminary analysis on the data itself.
This vignette walks through that preliminary analysis, then fits the full model:

1) Perform non-spatial exploratory item factor analysis to suggest a discrimination
   structure using the **mirt** package.
2) Analyse the empirical variogram of that fit's factor scores to suggest a spatial range
   using the **gstat** package.
3) Fit the full spatial model with `spifa()`, using the restriction and range prior from
   the previous two steps.

```{r}
library(mirt)
library(gstat)
library(sf)
library(spifa)
library(ggplot2)

data(ipixuna)
nitems <- ncol(ipixuna$items)
nfactors <- 3
```

## Step 1: Discrimination structure with `mirt`

`mirt()` requires named item columns; `ipixuna$items` doesn't have any, so we name
them here:

```{r}
items <- unclass(ipixuna$items)
colnames(items) <- paste0("item", seq_len(nitems))
head(items)
```

We fit an exploratory (unrestricted) 2PL item factor model and rotate it (varimax)
towards a simple structure:

```{r}
set.seed(11)
efa <- mirt(items, model = nfactors, itemtype = "2PL", verbose = FALSE)
loadings <- summary(efa, rotate = "varimax", verbose = FALSE)$rotF
loadings
```

Thresholding small loadings to zero turns this into a `0`/`1` restriction:

```{r}
A <- (abs(loadings) > 0.24) * 1
A
```

This is only one approach, different rotations could be applied depending on the
questionnaire and the latent construct under analysis. In this example, we will use `A` to
constrain discrimination parameters to zero.

## Step 2: Spatial range with `gstat`

The second thing `vignette("spifa")` takes as given is `priors$range`, the prior for each
Gaussian process's spatial range. `mirt`'s factor scores give a first estimate of each
household's latent abilities, which we can examine spatially with an empirical variogram.

### Estimate the latent factors

First, we obtain the estimated latent factors using `fscores` from **mirt**:

```{r}
fscore <- fscores(efa, rotate = "varimax")
colnames(fscore) <- paste0("F", seq_len(nfactors))
head(fscore)
```

Then, we need to geo-reference these factor scores in order to perform spatial analysis.
`gstat` uses the default coordinates units to compute distances between locations, so we
will also transform the coordinates system to a planar one where distances can be computed
in meters:

```{r}
fscore_sf <- st_sf(
  id = ipixuna$id, wealth = ipixuna$wealth, as.data.frame(fscore),
  geometry = st_transform(st_geometry(ipixuna), 32719)
)
fscore_sf
```

### Evaluate predictors effects

We can evaluate the effect of wealth on this estimated latent factors:

```{r, fig.height = 3}
fscore_sf |>
  tidyr::pivot_longer(F1:F3) |>
  ggplot(aes(wealth, value)) +
    geom_point() +
    geom_smooth(method = 'lm', formula = y ~ x) +
    facet_wrap(~ name) +
    theme_bw()
```

There seems to be an effect in F1 and F3, so we will analyse the residual spatial
structure after taking into account `wealth` effect:

### Evaluate the spatial structure

From `gstat`, we can use `variogram()` to obtain the empirical variogram, `vgm()` to define
a theoretical variogram and  `fit.variogram()` to fit the theoretical model to the
empirical variogram.

```{r, fig.height = 5}
vg1 <- variogram(F1 ~ wealth, fscore_sf, width = 20, cutoff = 1000)
vg1_fit <- fit.variogram(vg1, model = vgm(psill = 0.6, "Exp", range = 100, nugget = 0.1))
vg1_fit
plot(vg1, vg1_fit, pch = 19, main = "F1")
```

For F1, we can see that there is some evidence of spatial correlation at small scale.

```{r, fig.height = 5}
vg2 <- variogram(F2 ~ wealth, fscore_sf, width = 50, cutoff = 1000)
vg2_fit <- fit.variogram(vg2, model = vgm(psill = 0.6, "Exp", range = 100, nugget = 0.1))
vg2_fit
plot(vg2, vg2_fit, pch = 19, main = "F2")
```

For F2, there is a much clearer spatial structure with a range parameter around `170`.

```{r, fig.height = 5}
vg3 <- variogram(F3 ~ wealth, fscore_sf, width = 50, cutoff = 800)
vg3_fit <- fit.variogram(vg3, model = vgm(psill = 0.6, "Exp", range = 100, nugget = 0.1))
vg3_fit
plot(vg3, vg3_fit, pch = 19, main = "F3")
```

For F3 also the range parameter seems to be small.

In general, there is some evidence of spatial correlation in the latent factors, and
specifically in F2 the spatial structure is strong with a range parameter around 170. This
gives a good understanding of plausible values for the range parameters of our `spifa`
model. However, remember that the `spifa` and `mirt` models are not exactly the same, so
these values should be taken only as a reference to decide on our `constraints` and
`priors`.

## Step 3: Spatial modelling with `spifa`

With `A` from Step 1 and a range prior from Step 2 in hand, `spifa()` can fit the full
spatial model, one independent Gaussian process per factor by default. For the range
prior, F2's estimate is the one clear signal -- with only 100 households, F1 and F3's
much smaller ranges are more likely noise than genuine short-range structure -- so we
use a common reference value of 150 (close to F2's ~170) for all three factors:

```{r}
set.seed(1)
samples <- spifa(
  items ~ wealth, data = ipixuna, nfactors = nfactors,
  burnin = 1000, niter = 4000,
  constraints = list(discrimination = A, loading = diag(nfactors)),
  priors = list(range = list(initial = 150, mean = 150, sd = 0.4)))
samples
```

Note this uses `spifa()`'s own default (non-informed) initial values for everything
except the spatial range -- the same starting point a real user, without further
domain knowledge, would use. The full analysis can now proceed using the methods
presented in `vignette("spifa")` and the package's other vignettes.

### Visualize

Let's visualize the chains and densities of the posterior. Observe which parameters have
better mixing than others.

```{r, fig.height = 6}
plot(samples, c("c", "A"))
plot(samples, select = c("B", "T", "Corr"))
plot(samples, select = c("phi"))
```

### Predict

Let's visualize the spatial pattern of the latent factors mean prediction.

```{r}
pred_samples <- predict(samples)
plot_predict(pred_samples) +
  scale_colour_distiller(palette = "RdYlBu")
```

You can continue your analysis using the other methods and functions shown in the
package's vignettes.
